Reproducible Research Week 2 Project
====================================
```{r}
library(ggplot2)

library(dplyr)

if(!file.exists("./data"))
  {
    dir.create("./data")
  }
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl,destfile="./data/activity.zip")

unzip(zipfile="./data/activity.zip",exdir="./data")
activity <- read.csv("./data/activity.csv")
activity$date <- as.Date(activity$date)

head(activity)
```


#total number of steps taken per day
```{r}
stepsPerDay <- activity %>%
  group_by(date) %>%
    summarize(sumsteps = sum(steps, na.rm = TRUE)) 
```

#Display of first 10 rows of data
```{r}
head(stepsPerDay,10)
```


#histogram of the total number of steps taken each day
```{r}
hist(stepsPerDay$sumsteps, main="Histogram of Daily Steps",col="blue", xlab="Steps", ylim = c(0,30))
```

#mean and median of the total number of steps taken per day
```{r}
meanPreNA <- round(mean(stepsPerDay$sumsteps),digits = 2)
medianPreNA <- round(median(stepsPerDay$sumsteps),digits = 2)
print(paste("The mean is: ", meanPreNA))
print(paste("The median is: ", medianPreNA))
```

#average daily activity pattern
#time series plot (i.e. type = “l”|) of the 5-minute interval (x-axis) and the average number of steps taken
```{r}
stepsPerInterval <- activity %>%
  group_by(interval) %>%
    summarize(meansteps = mean(steps, na.rm = TRUE)) 
```
    
#Display first 10 rows of data
```{r}
head(stepsPerInterval,10)
plot(stepsPerInterval$meansteps ~ stepsPerInterval$interval,
     col="blue", type="l", xlab = "5 Minute Intervals", ylab = "Average Number of Steps",
     main = "Steps By Time Interval")
```

# 5-minute interval, on average across all the days in the dataset, containing the maximum number of steps
```{r}
print(paste("Interval containing the most steps on average:  ",stepsPerInterval$interval[which.max(stepsPerInterval$meansteps)]))
print(paste("Average steps for that interval: ",round(max(stepsPerInterval$meansteps),digits=2)))
```

#Imputing missing values
#total number of missing values in the dataset 
```{r}
print(paste("The total number of rows with NA is: ",sum(is.na(activity$steps))))
```

#strategy for filling in all of the missing values in the dataset
```{r}
head(activity,10)
activityNoNA <- activity  
  for (i in 1:nrow(activity))
    {
      if(is.na(activity$steps[i]))
        {
          activityNoNA$steps[i]<- stepsPerInterval$meansteps[activityNoNA$interval[i] ==  stepsPerInterval$interval]
        }
    }
head(activityNoNA,10)
```

#steps taken each day
```{r}
stepsPerDay <- activityNoNA %>%
  group_by(date) %>%
    summarize(sumsteps = sum(steps, na.rm = TRUE)) 
    head(stepsPerDay,10)

hist(stepsPerDay$sumsteps, main = "Histogram of Daily Steps", col="blue", xlab="Steps")
meanPostNA <- round(mean(stepsPerDay$sumsteps), digits = 2)
medianPostNA <- round(median(stepsPerDay$sumsteps), digits = 2)

print(paste("The mean is: ", mean(meanPostNA)))
print(paste("The median is: ", median(medianPostNA)))

NACompare <- data.frame(mean = c(meanPreNA,meanPostNA),median = c(medianPreNA,medianPostNA))
rownames(NACompare) <- c("Pre NA Transformation", "Post NA Transformation")
print(NACompare)
```


#differences in activity patterns between weekdays and weekends
#Create a new factor variable in the dataset
```{r}
activityDoW <- activityNoNA
activityDoW$date <- as.Date(activityDoW$date)
activityDoW$day <- ifelse(weekdays(activityDoW$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
activityDoW$day <- as.factor(activityDoW$day)
```

#panel plot containing a time series plot 
```{r}
activityWeekday <- filter(activityDoW, activityDoW$day == "weekday")
activityWeekend <- filter(activityDoW, activityDoW$day == "weekend")

activityWeekday <- activityWeekday %>%
  group_by(interval) %>%
    summarize(steps = mean(steps)) 
    activityWeekday$day <- "weekday"

activityWeekend <- activityWeekend %>%
  group_by(interval) %>%
    summarize(steps = mean(steps)) 
    activityWeekend$day <- "weekend"

wkdayWkend <- rbind(activityWeekday, activityWeekend)
wkdayWkend$day <- as.factor(wkdayWkend$day)


g <- ggplot (wkdayWkend, aes (interval, steps))
  g + geom_line() + facet_grid (day~.) + 
  theme(axis.text = element_text(size = 12),axis.title = element_text(size = 14)) + 
  labs(y = "Number of Steps") + labs(x = "Interval") + 
  ggtitle("Average Number of Steps - Weekday vs. Weekend") + 
  theme(plot.title = element_text(hjust = 0.5))
  ``` 